---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rCBN

<!-- badges: start -->
<!-- badges: end -->

The goal of rCBN is to provide an R interface to the CBN family of functions developed at ETH-Zurich.

## Installation

You can install the development version of rCBN like so:

``` r
devtools::install_github("rockwillck/rCBN")
```

## Example

rCBN provides an interface to both CT-CBN and H-CBN:

```{r example}
library(rCBN)

bc = Spock$new(
  poset = read_poset("../PATH/TO/examples/BC")$sets,
  numMutations = read_poset("../PATH/TO/examples/BC")$mutations,
  patternOrLambda = read_pattern("../PATH/TO/examples/BC")
)
ctcbn(bc)

hcbn(bc)
```

## CT-CBN
This project is based on and draws on code from CT-CBN from ETH Zurich.
CT-CBN. (n.d.). Department of Biosystems Science and Engineering – Department of Biosystems Science and Engineering | ETH Zurich. https://bsse.ethz.ch/cbg/software/ct-cbn.html