---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rCBN

<!-- badges: start -->
<!-- badges: end -->

The goal of rCBN is to provide an R interface to the CBN family of functions developed at ETH-Zurich.

## Installation

You can install the development version of rCBN like so:

``` r
devtools::install_github("rockwillck/rCBN")
```

## Example

rCBN provides an interface to CT-CBN, H-CBN, and B-CBN:

```{r example}
library(rCBN)
bc = Spock$new(
  poset = read_poset("inst/extdata/BC")$sets,
  numMutations = read_poset("inst/extdata/BC")$mutations,
  patternOrLambda = read_pattern("inst/extdata/BC")
)
```

```{r exampleCTCBN}
ctcbn(bc)
```

```{r exampleHCBN}
hcbn(bc)
```

```{r exampleBCBN, eval=FALSE}
bcbn(
  generateData(
    transitiveClosure(bc$poset),
    c(0.8, 0.7, 0.6, 0.7, 0.4, 0.25, 0.6, 0.75, 0.5, 0.2),
    0.1,
    400),
  n_cores = 4
  )
```

## CT-CBN
This project is based on and draws on code from CT-CBN from ETH Zurich.  
CT-CBN. (n.d.). Department of Biosystems Science and Engineering – Department of Biosystems Science and Engineering | ETH Zurich. https://bsse.ethz.ch/cbg/software/ct-cbn.html

## B-CBN
This project is based on and draws on code from B-CBN from ETH Zurich.  
bcbn. (n.d.). Department of Biosystems Science and Engineering – Department of Biosystems Science and Engineering | ETH Zurich. https://bsse.ethz.ch/cbg/software/bcbn.html