---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# CBN2Path

<!-- badges: start -->

<!-- badges: end -->

The goal of CBN2Path is to provide an R interface to the CBN family of functions developed at ETH-Zurich.

## Dependencies

`gsl` can be installed using:

``` bash
brew install gsl
```

If `gsl` was installed using any method other Homebrew, uninstall `gsl` before reinstalling using Homebrew.

If Homebrew is not installed, follow the instructions at <https://brew.sh> to install.

Make sure to restart R before proceeding.

## Installation

You can install the development version of `CBN2Path` like so:

``` r
devtools::install_github("rockwillck/CBN2Path")
```

## Example

`CBN2Path` provides an interface to CT-CBN, H-CBN, and B-CBN:

```{r example}
library(CBN2Path)
bc <- Spock$new(
    poset = read_poset("inst/extdata/BC")$sets,
    numMutations = read_poset("inst/extdata/BC")$mutations,
    genotypeMatrix = read_pattern("inst/extdata/BC")
)
```

```{r exampleCTCBN}
ctcbn(bc)
```

```{r exampleHCBN}
hcbn(bc)
```

```{r exampleBCBN}
bcbn()
```

## CT-CBN

This project is based on and draws on code from CT-CBN from ETH Zurich.\
CT-CBN. (n.d.). Department of Biosystems Science and Engineering – Department of Biosystems Science and Engineering \| ETH Zurich. <https://bsse.ethz.ch/cbg/software/ct-cbn.html>

## B-CBN

This project is based on and draws on code from B-CBN from ETH Zurich.\
bcbn. (n.d.). Department of Biosystems Science and Engineering – Department of Biosystems Science and Engineering \| ETH Zurich. <https://bsse.ethz.ch/cbg/software/bcbn.html>
